Extended Early Adulthood and Attitude Change

### 2024-06-18 manuscript/formative_period_kk.Rmd

For reasons explained below, we identified questions that were asked in at least three different panels with at least one of those panels asking some respondents to answer the question at three or more separate time points. Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14), we did not treat these as separate panels for the purposes of generating three distinct time points. In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis.

@@

When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's ability to report stable attitudes over time or to the structural features of the survey administration. Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change. Across our panels, duration between questions ranges from a few days (for questions asked in both pre- and post-election surveys in the ANES) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.

In the one-item, one-panel case (e.g., partisan identification in the 1972-76 panel), we could model the absolute change from wave-to-wave as a function of the duration between observations.

@@

If respondents have a stable underlying partisan identification they report with error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$. In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise). Conversely, $\beta_1$ captures the expected expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$). For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3, and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.

Our theoretical question deals primarily with whether these coefficients ($\beta_0$ and $\beta_1$) change over time. To explore that, we can add an additional term for time and interact the duration term with time.

@@

In this formulation, $\beta_2$ captures the change in the baseline level of random change over time. A negative value of $\beta_2$ would suggest that people are becoming more consistent reporters of their true underlying opinion over time. $\beta_3$ captures the expected change in absolute change over time for a one-year change in duration. A positive $\beta_3$ value would suggest that people are more likely to undergo a "real" change in their opinion over time in later panels than earlier panels.

@@

@@

The intercept suggests that, on average, respondents varied by about a fifth of the scale (17 points), controlling for duration. This is a rather high level of variance, but consistent with previous studies that find that, on any particular question, survey respondents tend to make large wave-to-wave changes. Variance around this intercept is quite large, with a standard deviation of about 5.67, meaning many items vary by as much as 20 percent of the scale, while some vary by less than 10 points.

Second, duration is weakly positively associated with wave-to-wave absolute change. For every additional year between waves, respondents' absolute deviance increases about .71 points. Notably, variance around this main effect is relatively small (.43), meaning that two thirds of questions have duration effects that fall between .23 and 1.14. Broadly speaking, we can say that most wave-to-wave change is non-persisting, consistent with other recent work [@kiley2020; @lersch2023]. 
The main effect of year of observation in decades is also small (-.37), but negative. In other words, on average, respondents exhibit less wave-to-wave change in their responses the longer a question is asked. However, there is significantly more variance around this main effect, with a standard deviation of about .88. This suggests that while some items show decreased wave-to-wave change over time, others show increased wave-to-wave change over time.

>>>>

For reasons explained below, we identified questions that were asked in at least three different panels with at least one of those panels asking some respondents to answer the question at three or more separate time points. Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14) and contain the same set of questions, we did not treat these as separate panels for the purposes of generating three distinct time points. In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis.

@@

When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's ability to report stable attitudes over time or to the structural features of the survey administration. Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change. Panel duration varied from less than one year (the 1980 ANES Major Panel Study) to five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.

In the one-item, one-panel case (e.g., partisan identification in the 1972-76 panel), we could model the absolute change from wave-to-wave as a function of an intercept and the duration between observations.

@@

If respondents have a stable underlying partisan identification they report with random error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$. In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise). Conversely, $\beta_1$ captures the expected expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$). For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3 and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations. More than two observations of the same question in a single panel produces variation in duration, which is why we ensure that each included question has at least one panel with three observations.

Our theoretical question deals primarily with the amount of change over time ($\beta_0$ and $\beta_1$) changes over time. To explore that, we can add an additional term for time and interact the duration term with time.

@@

In this formulation, $\beta_2$ captures the change in the baseline level of random change over time. A negative value of $\beta_2$ would suggest that people are becoming more consistent reporters of their true underlying opinion over time. $\beta_3$ captures the expected change in absolute change over time for a one-year change in duration. A positive $\beta_3$ value would suggest that people are more likely to undergo a durable or persisting change in their opinion over time in later panels than earlier panels.

@@

# Weights

We focus on using weights that accounts for sampling design.

For the 2016-20 ANES panel, almost all questions are asked in either the pre- or post-election version of both surveys. The panel includes weights for . The single item that was asked in the pre- and post-election waves of the 2016 panel (polviews) is weighted using the pre- election weights.

For the 2020-22 ANES panel.

@@

The intercept term in Model 1 suggests that, on average, respondents varied by about a fifth of the scale (17 points), controlling for duration, on any particular question. This is a rather high level of variance, but consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965]. At the same time, random-effect variance around this main intercept term is quite large, with a standard deviation of about 5.67. Many items vary by as much as 20 percent of the scale, while some vary by less than 10 points.

Second, duration between observations is positively but weakly associated with wave-to-wave absolute change. For every additional year between waves, respondents' absolute deviance increases about .71 points. Notably, variance around this main effect is relatively small (.43), meaning that two thirds of questions have duration effects that fall between .23 and 1.14. Given the comparatively high level of wave-to-wave change independent of duration, our results broadly suggest that most wave-to-wave change is non-persisting, consistent with other recent work [@kiley2020; @lersch2023].

The main effect of year of observation in decades is negative and also small (-.37). On average, respondents exhibit less wave-to-wave change in their responses in more recent panels. However, question-level random effects vary considerably around this main effect, with a standard deviation of about .88. This suggests that while some items show decreased wave-to-wave change over time, others show increased wave-to-wave change over time.
